---
title: "p8105_hw2_xl3371"
author: "Xuer Liu"
date: "2023-10-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

### First, clean the data in pols-month.csv
```{r, message=FALSE}
pols_month_clean = read_csv("./data/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(year = as.integer(year),
         day = as.integer(day)) %>%
  mutate(month = case_match(
    month, 
    "01" ~ "Jan",
    "02" ~ "Feb",
    "03" ~ "Mar",
    "04" ~ "Apr",
    "05" ~ "May",
    "06" ~ "Jun",
    "07" ~ "Jul",
    "08" ~ "Aug",
    "09" ~ "Sep",
    "10" ~ "Oct",
    "11" ~ "Nov",
    "12" ~ "Dec"
  )) %>%
  mutate(
    president = ifelse(prez_dem==1, "dem", "gop")) %>%
  select(-c(prez_gop, prez_dem, day))
```

### Second, clean the data in snp.csv
```{r}
snp_clean = read.csv("./data/snp.csv") %>%
  janitor::clean_names() %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  separate(date, into = c("year", "month", "day"), sep = "-") %>%
  mutate(year = as.integer(year),
         year = ifelse(year > 2049, year - 100, year),
    month = case_match(month, 
    "01" ~ "Jan",
    "02" ~ "Feb",
    "03" ~ "Mar",
    "04" ~ "Apr",
    "05" ~ "May",
    "06" ~ "Jun",
    "07" ~ "Jul",
    "08" ~ "Aug",
    "09" ~ "Sep",
    "10" ~ "Oct",
    "11" ~ "Nov",
    "12" ~ "Dec"
  )) %>%
  select(year, month, close) %>%
  arrange(year, month) %>%
  relocate(year, month)
```

### Third, tidy the unemployment data
```{r, message=FALSE}
unemployment_clean = read.csv("./data/unemployment.csv") %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "umemployment_rate"
  ) %>%
  rename(year = Year) %>%
  mutate(year = as.integer(year))
```

### Merge three datasets into result
```{r}
pols_snp = left_join(pols_month_clean, snp_clean, by = c("year","month"))
result = left_join(pols_snp, unemployment_clean, by = c("year","month"))
```

### Description
1. The `pols_month_clean` dataset contains `r ncol(pols_month_clean)` variables and `r nrow(pols_month_clean)` observations. The key variables include year, month, and president (elected president's party). This dataset tells us about the party affiliation distribution (democrat or republican) for governors and senators from years `r range(pols_month_clean$year)[1]` to `r range(pols_month_clean$year)[2]`. 
2. The `snp_clean` dataset contains `r ncol(snp_clean)` variables and `r nrow(snp_clean)` observations, ranging from `r range(snp_clean$year)[1]` to `r range(snp_clean$year)[2]`. The key variables include year, month, and close (closing values of stock index).
3. The `unemployment_clean` dataset contains `r ncol(unemployment_clean)` variables and `r nrow(unemployment_clean)` observations, ranging from `r range(unemployment_clean$year)[1]` to `r range(unemployment_clean$year)[2]`. The key variables include year, month, and unemployment rate.
4. The `result` dataset merges the above three datasets, with `r ncol(result)` columns and `r nrow(result)` rows. The data ranges from `r range(result$year)[1]` to `r range(result$year)[2]`, indicating the potential correlation between politics, closing value of stock, and unemployment.

## Problem 2

### Import and clean data from Mr. Trash Wheel sheet
```{r}
mr_tw = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N549") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(year = as.integer(year),
         type = "Mr",
         homes_powered = (weight_tons * 500) / 30) 
  
```

### Import and clean data from Professor Trash Wheel sheet
```{r}
prof_tw =
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = 2, range = "A2:M96") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(year = as.integer(year),
         type = "Professor",
         homes_powered = (weight_tons * 500) / 30) 
```

### Import and clean data from Gwynnda Trash Wheel sheet
```{r}
gw_tw =
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = 4, range = "A2:K108") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(year = as.integer(year),
         type = "Gwynnda",
         homes_powered = (weight_tons * 500) / 30) 
```

### Combine
```{r}
trashwheel_tidy = 
  bind_rows(mr_tw, prof_tw, gw_tw) %>% 
  janitor::clean_names() %>% 
  select(type, dumpster, everything()) %>% 
  arrange(type, dumpster)
```

### Description
The number of observations in `mr_tw` dataset is `r nrow(mr_tw)`. The key variables include `dumpster`, `weight_tons`, and `homes_powered`.
The number of observations in `prof_tw` dataset is `r nrow(prof_tw)`. The key variables include `dumpster`, `weight_tons`, and `homes_powered`.
The number of observations in `gw_tw` dataset is `r nrow(gw_tw)`. The key variables include `cigarette_butts`, `weight_tons`, and `homes_powered`.

The total weight of trash collected by Professor Trash Wheel is `r sum(filter(trashwheel_tidy, type=="Professor")$weight_tons)`

The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum((filter(trashwheel_tidy, type == "Gwynnda", year == 2021, month == "July"))$cigarette_butts)`.

