---
title: "p8105_hw2_xl3371"
author: "Xuer Liu"
date: "2023-10-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

### First, clean the data in pols-month.csv
```{r, message=FALSE}
pols_month_clean = read_csv("./data/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(year = as.integer(year),
         day = as.integer(day)) %>%
  mutate(month = case_match(
    month, 
    "01" ~ "Jan",
    "02" ~ "Feb",
    "03" ~ "Mar",
    "04" ~ "Apr",
    "05" ~ "May",
    "06" ~ "Jun",
    "07" ~ "Jul",
    "08" ~ "Aug",
    "09" ~ "Sep",
    "10" ~ "Oct",
    "11" ~ "Nov",
    "12" ~ "Dec"
  )) %>%
  mutate(
    president = ifelse(prez_dem==1, "dem", "gop")) %>%
  select(-c(prez_gop, prez_dem, day))
```

### Second, clean the data in snp.csv
```{r}
snp_clean = read.csv("./data/snp.csv") %>%
  janitor::clean_names() %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  separate(date, into = c("year", "month", "day"), sep = "-") %>%
  mutate(year = as.integer(year),
         year = ifelse(year > 2049, year - 100, year),
    month = case_match(month, 
    "01" ~ "Jan",
    "02" ~ "Feb",
    "03" ~ "Mar",
    "04" ~ "Apr",
    "05" ~ "May",
    "06" ~ "Jun",
    "07" ~ "Jul",
    "08" ~ "Aug",
    "09" ~ "Sep",
    "10" ~ "Oct",
    "11" ~ "Nov",
    "12" ~ "Dec"
  )) %>%
  select(year, month, close) %>%
  arrange(year, month) %>%
  relocate(year, month)
```

### Third, tidy the unemployment data
```{r, message=FALSE}
unemployment_clean = read.csv("./data/unemployment.csv") %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "umemployment_rate"
  ) %>%
  rename(year = Year) %>%
  mutate(year = as.integer(year))
```

### Merge three datasets into result
```{r}
pols_snp = left_join(pols_month_clean, snp_clean, by = c("year","month"))
result = left_join(pols_snp, unemployment_clean, by = c("year","month"))
```

### Description
1. The `pols_month_clean` dataset contains `r ncol(pols_month_clean)` variables and `r nrow(pols_month_clean)` observations. The key variables include year, month, and president (elected president's party). This dataset tells us about the party affiliation distribution (democrat or republican) for governors and senators from years `r range(pols_month_clean$year)[1]` to `r range(pols_month_clean$year)[2]`. 
2. The `snp_clean` dataset contains `r ncol(snp_clean)` variables and `r nrow(snp_clean)` observations, ranging from `r range(snp_clean$year)[1]` to `r range(snp_clean$year)[2]`. The key variables include year, month, and close (closing values of stock index).
3. The `unemployment_clean` dataset contains `r ncol(unemployment_clean)` variables and `r nrow(unemployment_clean)` observations, ranging from `r range(unemployment_clean$year)[1]` to `r range(unemployment_clean$year)[2]`. The key variables include year, month, and unemployment rate.
4. The `result` dataset merges the above three datasets, with `r ncol(result)` columns and `r nrow(result)` rows. The data ranges from `r range(result$year)[1]` to `r range(result$year)[2]`, indicating the potential correlation between politics, closing value of stock, and unemployment.

## Problem 2

